{\bf \Large The Power of Context in Decision-Making and Recommendations}

### 2023-07-03 draft.tex
Definite draft

Firstly, I extend my deepest gratitude to Zakaria Babutsidze, my PhD supervisor. His pledge to transition my perspective from engineering to social science laid the foundation for this journey. I am indebted to Francesco Castellanetta, our PhD program director, for his guiding hand and unwavering faith in my capabilities. To Ludovic Diabaggio, I appreciate our intriguing discussions that transcended academia. I am grateful to William Rand for his invaluable insights and co-authorship. Daniela Iubatti, your informal talks and unwavering support have been a source of inspiration. I appreciate Diego Zunino for his earnest and invaluable guidance. Bruno Cirillo deserves special mention; our albeit slightly formal discussions have significantly shaped my work. My heartfelt appreciation goes to Renata Kaminska. Her support and intuitive understanding of my thoughts were indispensable in my pursuits. I am grateful to Benjamin Montmartin; our separate exchanges about wine (we were in France in the end) and statistics enriched my perspectives immeasurably. To Valerio Incerti, our creative discourses, even about less creative ideas were a constant source of inspiration. Lapo Mola's invaluable input became a source of aspiration, for which I am grateful.

Aytan, my beloved wife and best friend, your unwavering support and guidance have been my beacon during tumultuous times. Ecem Delicik, my dear friend, I am grateful for the shared stories, laughter, and silence that have strengthened our friendship. The moment you realized I was speaking to you in Turkish is unforgettable. Johanna Deperi, our shared moments and long talks were invaluable for self-reflection, joy, and sometimes, ``justified sadness'' throughout this journey. Artyom Yepremyan, our discussions helped us conquer the PhD mountain. I will forever cherish the times when I needed a hand, and you extended yours. Manon Desjardins, your constant support despite the towering stack of papers on your desk will always be appreciated. Mehdi Ibrahim, I am grateful for your steady guidance and the irresistible desserts from Morocco. Teymur Mardaliyev, your friendship and coding prowess have been invaluable; only you could convince me to use JavaScript there.

Finally, no amount of words can fully express my gratitude to my parents. Dad, your inquisitive nature has been my guiding star, instilling in me the curiosity that set me on this scholarly path. Your enthusiasm for the uncharted has been infectious, leading me to question, explore, and discover. Mom, you have been my anchor, reminding me of my roots and my identity amidst the academic whirlwind. Your unwavering encouragement and steadfast belief have reinforced my sense of self, allowing me to face challenges with courage and resilience. You taught me to stay grounded and true to myself, no matter how far I travel in the pursuit of knowledge. Your lessons continue to shape my journey, helping me navigate both personal and academic hurdles with grace and fortitude. Çox sağ olun ki, varsınız!

@@

Recommender systems have become a crucial ally in the vast landscape of digital world. They help individuals to navigate through the plethora of alternatives and find what they are looking for. They achieve this by using sophisticated algorithms that analyze the wide space of items, users and their interactions with each other. System designers and businesses thrive to maximize the accuracy of recommendations, meaning they want to maximize the consumption of the recommendations. However, recent research stream shows that, accuracy is not the only way to make recommender systems effective. Other factors like diversity, serendipity \citep{kaminskas2016diversity}. Moreover, the black-box nature of the algorithms do not allow to understand the reasons behind a particular user's behaviour \citep{kotkovSurveySerendipityRecommender2016, samih2021exmrec2vec}. Hence, more and more interest is driven towards the understanding and utilizing the choice context \citep{adomavicius2005toward}. Choice context has received interest in scenarios where user or item information is scarce \citep{scheinMethodsMetricsColdStart}.

Choice context has long received the attention of scholars in many fields, including marketing, psychology, economics. Because it was studied in many domains, scholars refer to it differently. This thesis concentrates and uses two of them. The first one posits that choice context \footnote{For the remainder of the dissertation, I will follow the previous literature \citep{truebloodEtAl13} and refer to this definition of choice context as context effects.} is ``the availability and the nature of choice alternatives'' \citep{tversky1972elimination, huberEtAl82, simonson89}. Previous research has demonstrated the existence of context effects in various settings \citep{herne1997decoy, soltani2012range, evangelidisEtAl18,  wuConsguner20}. The second definition of context arises much later with the proliferation of recommender systems. Context also referred to as ``the time and content of the choice, the location or socio-demographic characteristics of the decision-maker...'' \citep{adomavicius2011context}\footnote{To distinguish between these two definitions I will refer to this definition as external context throughout this dissertation.}

However, modeling for these effects mathematically was a challenge because the existing models suffered from independence of irrelevant alternatives criterion \citep{luce59}, meaning these models treated each alternative in isolation. This has shifted the scholar's interest towards computational models \citep{usher2001time, roe2001multialternative, trueblood2014multiattribute, noguchi2018multialternative}. Yet, none of these models has been applied to real-world choice data and as a result, their applicability to field data remained an open question. I address this gap in the first study where I apply a computational model to a field data of high heterogeneity among dimensions. Building on my results in my second study I delve into the choice modeling more and develop a methodology to account for three main context effects studied in the literature \citep{truebloodEtAl13}.

When new users or items are introduced to recommender systems, they fail to function as intended because of the lack of information their algorithms need. In such cases, even when they do provide recommendations, recommendations are far from personalized \citep{lika2014facing}. It is considered to be a key challenge in recommender system design \citep{park2009pairwise}. Previous literature has applied various methods to overcome this problem \citep{guy2009personalized, aharon2013off, bykau2013coping, saveski2014item}. However, the limitations of these approaches are that they ignore the context effects and concentrate on scenarios where information scarcity is temporary. Using the findings from my second study, decision-making and choice modeling literature I propose an innovative solution to address continuous cold-start issue in recommender system design by utilizing contextual information of the menu to generate relevant choice sets through a two-step choice modeling method.

@@

    Previous computational decision making models developed to account for context effects have only been
    studied with an experimental data where only one effect was produced at a time. Using data coming from strictly controlled experimental environments
    hinders the understanding of context effects that occur in real-world choice scenarios where items have multiple dimensions and choice sets have dozens of alternatives. In this chapter
    I apply a computational model to an observational data which was not done before. The data comes from an air travel industry and is ideal to study context effects in multi-attribute, multia-lternative choice environments. I first find optimal parameters for computational model using differential evolution algorithm. Then, I complement a traditional choice model with its outputs and assess the significance of its contribution. This chapter contributes to context effect and decision making literature by providing further insights on behavior of computational decision making models in real-world choice data.

@@

In the last three decades, researchers have developed many computational models which account for context effects: Multialternative Decision Field Theory (MDFT) \citep{roe2001multialternative}, Leaky competing
accumulator \citep{usher2001time}, Multiattribute linear ballistic accumulator \citep{trueblood2014multiattribute}, Multialternative Decision by Sampling  \citep{noguchi2018multialternative}. Some of these models have been extensively tested and studied, while others are relatively new, hence have not received much attention from scholars \citep{truebloodEtAl13}. However, these studies have been performed with experimental data \citep{trueblood2014multiattribute,  berkowitsch2014rigorously, evans2019response, busemeyer2019cognitive}. Research has proven that the behavior of individuals in laboratory choice environment is different from real-world choice environment \citep{hogarth1989risk}. Hence, the applicability of such models to field data is an uninvestigated avenue, because no previous study has been done where a computational model was been applied to a real-world observational data. The plan is to address this gap by applying Multialternative Decision by Sampling (MDbS) model proposed by Noguchi and Stewart \citeyearonly{noguchi2018multialternative} to a field data from airfare booking domain. This would allow me to assess the applicability level of this model to multialternative and multidimensional field data.

@@

We make choices all the time. Remember the time when you went to see a movie you have been waiting for some time ago and decided to grab some popcorn before entering. You might have seen something like in the figure \ref{fig:decoyPopcornExample}  although prices may be higher these days. You are puzzled at first, but reminded that the movie about to start, so you better hurry up and choose one.

@@

However, research has concluded that the context of the choice set and options in it have substantial effect on how people make choices. This effect has been studied extensively for the past five decades by many economists, marketing scholars and psychologists \footnote{See Dowling et al. \citeyearonly{dowlingEtAl20} and Lichtenstein \citeyearonly{lichtenstein2006construction} for more comprehensive review.} \citep{ kahnemanTversky79, simonson89, tverskySimonson93, lichtenstein2006construction, dowlingEtAl20}. Most of the research has focused on three context effects, also reffered to as ``big three'': attraction, compromise and similarity \citep{howes2016contextual}.

@@

The very first computational model which could account for all three context effects was Multialternative Decision Field Theory (MDFT) developed by Roe et al. \citeyearonly{roe2001multialternative} as an extension of Decision Field Theory \citep{busemeyer1993decision}. It is a dynamic model of decision making that accommodates multialternative preferential choice situations, which was not possible with Decision Field Theory \citep{hotaling2019quantitative}. MDFT assumes that the decision making can be explained in three general mechanisms. First, attention allocation posits that attention switches over time between attributes stochastically. Second, evaluation mechanism posits that attribute value of given option is compared with the average attribute values of other options  which makes sure that each option in the choice set participates in comparison. Third, evidence accumulation mechanism which based on the result of evaluations, gathers evidence in favor of alternatives compared. As soon as gathered evidence hits the externally set relative threshold choice is concluded \citep{busemeyer2002survey}. This means, as soon as the difference between the highest and the second highest evidence values is larger than the relative threshold, choice is made. If this threshold is not met, the decision continues until the pre-set time limit is reached.

@@

Multiattribute linear ballistic accumulator (MLBA) is another attention based decision making model first proposed by Trueblood et al. \citeyearonly{trueblood2014multiattribute}. Similar to MDFT it is also a dynamic model which can be explained in three general mechanisms: attention allocation, evaluation of alternatives and evidence accumulation. However, the two models have key differences. Firstly, MDFT emulates the search process of elimination by aspects proposed by Tversky \citeyearonly{tversky1972elimination} assuming decision-makers compare alternatives to each other over time. In contrast, MLBA assumes that individuals make comparisons and accumulate evidence all alternatives indepentently from one-another at the same time and then accumulate evidence \citep{trueblood15fragile}. Also, MDFT assumes that individuals have limited cognitive capacity to process information when comparing items together in contrast to MLBA, which considers individuals with unlimited cognitive capacity. Moreover, unlike MDFT, where decision is made based on relative threshold, in MLBA decision is based of absolute threshold, i.e. as soon as one alternative's evidence reaches the threshold, decision is made in favor of that alternative. Another difference between these two models is the context effects they account for. While MDFT does account for attraction, compromise and similarity effects \citep{hotaling2019quantitative}, MLBA additionally accounts for preference reversals arising from context \citep{trueblood15fragile}.

@@

According to MDbS, when comparing two tickets between Paris and New-York, the price of a ticket would be compared to the prices of other tickets in the choice set and also to the ones and which an individual has previously seen, but are not in the current choice set. Comparisons are ordinal, meaning that evidence accumulated towards the ``winner'' at a rate of one irregardless of how large the difference was.

@@

\begin{align}\label{probabilityOneIsFavored}
    P(m_{ij} \text{ is favored over } m_{kj}) = 
        \begin{cases}
        F(\beta_0 (| \frac{m_{ij} - m_{kj}}{m_{kj}} |- \beta_1)) & \text{if } A_i > X_i \\
        0 & \text{otherwise}
        \end{cases} ,
\end{align}
where $F$ is a logistic sigmoid function and $\beta_0$ and $\beta_1$ corresponde to the advantage value and the probability that this particular advantage value will be enough to be preferred. For example, consider the case where $\beta_0 = 0.1$ and $\beta_1 = 50$. This would mean that the advantage of 10\% would be preferred with 50\% probability. Consequently, if the difference is 20\%, then it will be preferred with 99\% probability. The logistic function brings the notion of ``soft'' comparison instead of ``hard'' comparison in which case the small differences would be ignored, while large differences would be extremely preferred \citep{noguchi2018multialternative}.

@@

When Tversky \citeyearonly{tversky1972elimination} explained to similarity effect in a choice set consisting of $A$, $B$ and $S$ he explained it via his famous elimination by aspects theory. When $B$ and $S$ is similar to each other, they will either be eliminated together, or stay together. Hence, having $S$ in the choice set will steal the probability of choice from $B$.

@@

Not only MDbS accounts for big three context effect, it also successfully accounts for other known effects to decision making literature such as: attribute spacing effect \citep{cooke1998multiattribute}, centrality effect \citep{brown2011decision}, background contrast effect \citep{tverskySimonson93}, endowment effect \citep{knetsch1989endowment} and etc.\footnote{For the full list of the effects MDbS can account for please refer to Noguchi \citeyearonly{noguchi2018multialternative}.} Overall, authors claim that MDbS can theoretically account for up to 25 context effect variations.

@@

Because of MDbS nature of comparing dimensions with one another, for the purposes of this study, I can not use variables that do not differ across the menus. I can not utilize horizontal attributes in my analysis because they do not follow the standard ``greater the better'' mathematical approach to comparison either. However, these variables are used later in this dissertation in chapters \ref{chapter:jmrPaper} and \ref{chapter:hicssPaper}. Hence, I am bound to utilize solely four vertical attributes in my analysis. Table \ref{tab:descriptiveStats} provides descriptive information about variables. I have also multiplied all four vertical variables by \-1 to convert them to negative scale because of MDbS nature of comparing absolute values and the assumption that consumers prefer lower values along vertical dimensions.

@@

In the course of our study, I also introduce an additional dataset sourced from a controlled experiment conducted by Noguchi \citeyearonly{noguchi2018multialternative}, distinct from the primary observational data. Albeit not as diverse in terms of alternatives, dimensions and choice sets, this experimental data carries significant value, not as a principal analytic focus, but rather as a mean to corroborate my main findings. I will apply the same analytical techniques employed in the observational analysis, and utilize experimental dataset as a robustness check to verify the validity of my results. Henceforth, the role of this data is primarily confirmatory.

This data comes from an experiment conducted by Noguchi \citeyearonly{noguchi2018multialternative} where 503 participants, aged 18 to 75, which took part in Amazon Mechanical Turk resulting in 5295 observations total. Participants faced eight randomly sampled decision scenarios with descriptions that consisted of two and three alternative sets, each ternary choice set containing only one of the attraction, compromise, similarity effects. For ternary choice sets, to create a context effect one alternative was randomly selected as focus and third alternative was generated following three scenarios: a) for attraction effect scenario, third options' both dimensions were reduced by 25\% of the difference between the remaining two options' dimensions; b) for compromise scenario, the third option was generated in a way that it would make the randomly chosen target a compromise; c) in similarity scenario 2\% of the difference was added to one dimension while 2\% was subtracted from another dimension for the third option. The dimensions of alternatives are described in the table \ref{tab:noguchiDescriptions}.

@@

After discussing the importance of optimal parameter search and defining the optimization algorithm, the question remaining becomes the evaluation metric of the MDbS. Previous studies which have applied various dynamic choice models to experimental data, have used the mean absolute error of aggregate choice shares for the entire dataset as main metric. Albeit an interesting approach itself, this will not be a feasible approach for me because the experimental data these models have been applied to entailed ternary choice sets, whereas the observational data is not completely ternary. It comprised of choice sets with minimum of 2 and maximum of 99 alternatives.

Choice set designers and engineers have long used ``Top n'' accuracy metrics when designing choice sets, or testing the performance of statistical models \citep{ricci2015recommender}. ``Top $n$'' accuracy metric posits whether or not, true class of the option matches the top $n$ predictions of the model. I will follow and adopt this metric because it is well established in the literature, mirrors the real-world decision making and it fits the contribution of the thesis the best. I will use ``Top 1'' accuracy which ranges from 0 to 1 as my metric of optimization for an individual choice set. Because choice sets in the data vary significantly in size, I will use average Top 1 accuracy metric weighted by menu size.This will ensure that smaller menus contribute proportional to their sizes. Also, to comply with differential evolution algorithm's aim of minimization, I will multiply this metric by -1. As an additional measure to explore the parameter space thoroughly, I employ multiple-run approach for differential evolution algorithm. Specifically, I will execute it ten times across the entire dataset. This reposition will allow me to further explore the parameter space, mitigating the risk of missing any region that can potentially contain optimal solution.

@@

Recall that variable ``MDbS output'' refers to the probabilities produced by MDbS. Model 2 results show positive and statistical significance effect for the information provided. It shows that MDbS is able to capture additional information about the choice by accounting for context effects. To better understand the significance of this result, figure \ref{fig:marginsAmadeusGraph} shows the average marginal effects of the information provided by computational model. One can immediately observe the downward trend. It is not surprising. As the number of alternatives increases, each additional alternative adds less to the likelihood of choice than the one before. In the context of menus, it could imply that when there are fewer options (smaller menus), the likelihood of any particular choice being selected is more significantly influenced by MDbS output. On average, for every 0.5 increase in MDbS output the probability of choice has increased by 0.037 percentage points \footnote{Marginal effect of MDbS output across the whole dataset was 0.074.}. This effect was as high as 0.12 percentage points for menus containing as small as 5 alternatives.

@@

At first sight, the results from optimization indicate that the optimal parameters differ across the datasets. Considering the nature for these two datasets, such result is to be expected. While optimal $\alpha$ tends to be higher than the one in observational data, both optimal $\beta_0$ and $\beta_1$ values are lower than their counterparts. Higher $\alpha$ can indicate that for smaller choice sets, MDbS tends to less strict in comparison criteria. While the performance metrics seem higher than for the observational data, menus are considerably smaller. Table \ref{tab:optimizationNoghuchiResults} entails further information.

@@

Also, the use of MDbS in current study has shown it has ability to potentially capture wide range of context effects, including attraction, compromise and similarity. While these results have yielded valuable insights, the general nature of MDbS and other computational models is their inability to successfully isolate those effects from one another. Main reason for that is that they have been only tested in experiments with one effect present at a time. When the number of alternatives in the dataset increases the potential interplay between options and the existence of other context effects come into play.

My findings provide a strong foundation which leads to a crucial but also challenging future direction: the development of a methodology which would allow to disentangle this ``general'' context effect. I will computationally differentiate among three main components, attraction, similarity and compromise in a multidimensional, multialternative choice setting. This goal provides a great motivation for the next chapter of my thesis.

@@

Considering the rapid technological developments of the recent years, we may state with confidence that user control in recommender systems is about to take a huge leap forward and open even more research directions.

>>>>

Firstly, I extend my deepest gratitude to Zakaria Babutsidze, my PhD supervisor. He has been a great mentor and a source of inspiration for me. His honesty and straightforwardness challenged me to think critically and creatively, shaping me into someone better day by day. I am indebted to Francesco Castellanetta, our PhD program director, for his guiding hand and unwavering faith in my capabilities. I appreciate our long and intriguing discussions with Ludovic Diabaggio that transcended academia. I am grateful to William Rand for his invaluable insights and co-authorship. Daniela Iubatti, your informal talks and unwavering support have been a source of inspiration. I appreciate Diego Zunino for his earnest and invaluable guidance. Bruno Cirillo deserves special mention; our albeit slightly formal discussions have significantly shaped my work. My heartfelt appreciation goes to Renata Kaminska. Her support and intuitive understanding of my thoughts were indispensable in my pursuits. I am grateful to Benjamin Montmartin; our separate exchanges about wine (we were in France in the end) and statistics enriched my perspectives immeasurably. To Valerio Incerti, our creative discourses, even about less creative ideas were a constant source of inspiration. Lapo Mola’s precious feedback motivated me to pursue excellence, and I value that wholeheartedly.

Aytan, my beloved wife and best friend, your unwavering support and guidance have been my Polaris guiding me through the darkest nights and taught me to only settle for infinity and beyond. Ecem Delicik, my dear friend, I am grateful for the shared stories, moments, laughter, and silence that have strengthened our friendship. The moment you realized I was speaking to you in Turkish is unforgettable (I am sure, you still remember it too). Johanna Deperi, our shared moments and long talks were invaluable for self-reflection, joy, and sometimes, ``justified sadness'' throughout this journey. Artyom Yepremyan, our discussions helped us conquer the PhD mountain. I will forever cherish the times when I needed a hand, and you extended yours. Manon Desjardins, your constant support despite the towering stack of papers on your desk will always be appreciated. Mehdi Ibrahim, I am grateful for our coffee talks and the irresistible desserts from Morocco. Teymur Mardaliyev, your friendship and coding prowess have been invaluable; only you could convince me to settle for JavaScript there.

Finally, no amount of words can fully express my gratitude to my parents. Dad, your inquisitive nature has been my guiding star, instilling in me the curiosity that set me on this scholarly path. Your enthusiasm for the uncharted has been infectious, leading me to question, explore, and discover. Mom, you have been my anchor, reminding me of my roots and my identity amidst the academic whirlwind. Your unwavering encouragement and steadfast belief have reinforced my sense of self, allowing me to face challenges with courage and resilience. You taught me to stay grounded and true to myself, no matter how far I travel in the pursuit of knowledge. Your lessons continue to shape my journey, helping me navigate both personal and academic hurdles with grace and fortitude.

\textit{Çox sağ olun ki, varsınız!}

@@

Recommender systems have become a crucial ally in the vast landscape of digital world. They help individuals to navigate through the plethora of alternatives and find what they are looking for. They achieve this by using sophisticated algorithms that analyze the wide space of items, users and their interactions with each other. System designers and businesses thrive to maximize the accuracy of recommendations, meaning they want to maximize the consumption of the recommendations. However, recent research stream shows that, accuracy is not the only way to make recommender systems effective. Other factors like diversity, serendipity are important too\citep{kaminskas2016diversity}. Moreover, the black-box nature of the algorithms do not allow us to understand the reasons behind a particular user's choice behaviour \citep{kotkovSurveySerendipityRecommender2016, samih2021exmrec2vec}. Hence, more and more interest is driven towards the understanding and utilizing the context around the particular choice \citep{adomavicius2005toward}.

Choice context has long received the attention of scholars in many fields, including marketing, psychology, economics. Because it was studied in many domains, scholars refer to it differently. This thesis concentrates on and uses two of them. The first one posits that choice context \footnote{For the remainder of the dissertation, I will follow the previous literature \citep{truebloodEtAl13} and refer to this definition of choice context as context effects.} is ``the availability and the nature of choice alternatives'' \citep{tversky1972elimination, huberEtAl82, simonson89}. Previous research has demonstrated the existence of context effects in various settings \citep{herne1997decoy, soltani2012range, evangelidisEtAl18,  wuConsguner20}. The second definition of context arises much later with the proliferation of recommender systems. Then, context is also referred to as ``the time and content of the choice, the location or socio-demographic characteristics of the decision-maker...'' \citep{adomavicius2011context}\footnote{To distinguish between these two definitions I will refer to this definition as external context throughout this dissertation.}

However, modeling for context effects mathematically was a challenge because the existing models suffered from independence of irrelevant alternatives criterion \citep{luce59}, meaning these models treated each alternative in isolation. This has shifted the scholar's interest towards computational models \citep{usher2001time, roe2001multialternative, trueblood2014multiattribute, noguchi2018multialternative}. Yet, none of these models has been applied to real-world choice data and as a result, their applicability to field data remained as an open question. I address this gap in the first study where I apply a computational model to a field data of high heterogeneity among dimensions. Building on my results in my second study I delve into the choice modeling more and develop a methodology to account for three main context effects studied in the literature \citep{truebloodEtAl13}.

When new users or items are introduced to recommender systems, they fail to function as intended because of the lack of information their algorithms need. In such cases, even when they do provide recommendations, recommendations are far from personalized \citep{lika2014facing}. It is considered to be a key challenge in recommender system design \citep{park2009pairwise}. Previous literature has applied various methods to overcome this problem, including asking users to rate some items, share their preferences among others \citep{guy2009personalized, aharon2013off, bykau2013coping, saveski2014item}. However, the limitations of these approaches are that they ignore the context effects and concentrate on scenarios where information scarcity is temporary. Using the findings from my second study and combining it with decision-making and choice modeling literature I propose an innovative solution to address continuous information scarcity issue in recommender system design by utilizing contextual information of the menu to generate relevant choice sets through a two-step choice modeling method.

@@

    Previous computational decision making models that were developed to account for context effects have only been
    studied with an experimental data where only one effect was produced at a time. Using data coming from strictly controlled experimental environments
    hinders the understanding of context effects that occur in real-world choice scenarios where items have multiple dimensions and choice sets have dozens of alternatives. In this chapter
    I apply a computational model that accounts for context effects to an observational data which was not done before. The data comes from an air travel industry and is ideal to study context effects in multi-attribute, multi-alternative choice environments. I first find optimal parameters for computational model using differential evolution algorithm. Then, I complement a traditional choice model with its outputs and assess the significance of its contribution. This chapter contributes to context effect and decision making literature by providing further insights on behavior of computational decision making models in real-world choice data.

@@

In the last three decades, researchers have developed many computational models which account for context effects: Multialternative Decision Field Theory (MDFT) \citep{roe2001multialternative}, Leaky competing
accumulator \citep{usher2001time}, Multiattribute linear ballistic accumulator \citep{trueblood2014multiattribute}, Multialternative Decision by Sampling  \citep{noguchi2018multialternative} are part of them. Some of these models have been extensively tested and studied, while others are relatively new, hence have not received much attention from scholars \citep{truebloodEtAl13}. However, these studies have been performed with experimental data \citep{trueblood2014multiattribute,  berkowitsch2014rigorously, evans2019response, busemeyer2019cognitive}. Research has proven that the behavior of individuals in laboratory choice environment is different from real-world choice environment \citep{hogarth1989risk}. Hence, the applicability of such models to field data is an uninvestigated avenue, because no previous study has been done where a computational model was been applied to a real-world observational data. The plan is to address this gap by applying Multialternative Decision by Sampling (MDbS) model proposed by Noguchi and Stewart \citeyearonly{noguchi2018multialternative} to a field data from airfare booking domain. This would allow me to assess the applicability level of this model complex field-data.

@@

We make choices all the time. Imagine the time when you went to see a movie you have been waiting for some time ago and decided to grab some popcorn before entering. You might have seen something like in the figure \ref{fig:decoyPopcornExample}, although prices may be higher these days. You are puzzled at first, but reminded that the movie about to start, so you better hurry up and choose one.

@@

However, research has concluded that the context of the choice set and options in it have substantial effect on how people make choices. This effect has been studied extensively for the past five decades by many economists, marketing scholars and psychologists \footnote{See Dowling et al. \citeyearonly{dowlingEtAl20} and Lichtenstein \citeyearonly{lichtenstein2006construction} for more comprehensive review.} \citep{ kahnemanTversky79, simonson89, tverskySimonson93, lichtenstein2006construction, dowlingEtAl20}. Most of the research has focused on three context effects, also referred to as ``big three'': attraction, compromise and similarity \citep{howes2016contextual}.

@@

The very first computational model which could account for all three context effects was Multialternative Decision Field Theory (MDFT) developed by Roe et al. \citeyearonly{roe2001multialternative} as an extension of Decision Field Theory \citep{busemeyer1993decision}. It is a dynamic model of decision making that accommodates multi-alternative preferential choice situations, which was not possible with Decision Field Theory \citep{hotaling2019quantitative}. MDFT assumes that the decision making can be explained in three general mechanisms. First, attention allocation posits that attention switches over time between attributes stochastically. Second, evaluation mechanism posits that attribute value of given option is compared with the average attribute values of other options  which makes sure that each option in the choice set participates in comparison. Third, evidence accumulation mechanism which based on the result of evaluations, gathers evidence in favor of alternatives compared. As soon as gathered evidence hits the externally set relative threshold, choice is concluded \citep{busemeyer2002survey}. This means, as soon as the difference between the highest and the second highest evidence values is larger than the relative threshold, choice is made. If this threshold is not met, the decision continues until the pre-set time limit is reached.

@@

Multiattribute linear ballistic accumulator (MLBA) is another attention based decision making model first proposed by Trueblood et al. \citeyearonly{trueblood2014multiattribute}. Similar to MDFT it is also a dynamic model which can be explained in three general mechanisms: attention allocation, evaluation of alternatives and evidence accumulation. However, the two models have key differences. Firstly, MDFT emulates the search process of elimination by aspects proposed by Tversky \citeyearonly{tversky1972elimination} assuming decision-makers compare alternatives to each other over time. In contrast, MLBA assumes that individuals make comparisons and accumulate evidence all alternatives independently from one-another at the same time and then accumulate evidence \citep{trueblood15fragile}. Also, MDFT assumes that individuals have limited cognitive capacity to process information when comparing items together in contrast to MLBA, which considers individuals with unlimited cognitive capacity. Moreover, unlike MDFT, where decision is made based on relative threshold, in MLBA decision is based of absolute threshold, i.e. as soon as one alternative's evidence reaches the threshold, decision is made in favor of that alternative. Another difference between these two models is the context effects they account for. While MDFT does account for attraction, compromise and similarity effects \citep{hotaling2019quantitative}, MLBA additionally accounts for preference reversals arising from context \citep{trueblood15fragile}.

@@

According to MDbS, when comparing two tickets between Paris and New-York for example, the price of a ticket would be compared to the prices of other tickets in the choice set and also to the ones and which an individual has previously seen, but are not in the current choice set. Comparisons are ordinal, meaning that evidence accumulated towards the ``winner'' at a rate of one irregardless of how large the difference was.

@@

\begin{align}\label{probabilityOneIsFavored}
    P(m_{ij} \text{ is favored over } m_{kj}) = 
        \begin{cases}
        F(\beta_0 (| \frac{m_{ij} - m_{kj}}{m_{kj}} |- \beta_1)) & \text{if } A_i > X_i \\
        0 & \text{otherwise}
        \end{cases} ,
\end{align}
where $F$ is a logistic sigmoid function and $\beta_0$ and $\beta_1$ correspond to the advantage value and the probability that this particular advantage value will be enough to be preferred. For example, consider the case where $\beta_0 = 0.1$ and $\beta_1 = 50$. This would mean that the advantage of 10\% would be preferred with 50\% probability. Consequently, if the difference is 20\%, then it will be preferred with 99\% probability. The logistic function brings the notion of ``soft'' comparison instead of ``hard'' comparison in which case the small differences would be ignored, while large differences would be extremely preferred \citep{noguchi2018multialternative}.

@@

When Tversky \citeyearonly{tversky1972elimination} explained to similarity effect in a choice set consisting of $A$, $B$ and $S$ he explained it via his famous elimination by aspects theory. When $B$ and $S$ is similar to each other, they will either be eliminated together, or stay together. Hence, having $S$ in the choice set will ``steal'' the probability of choice from $B$.

@@

Not only MDbS accounts for big three context effect, it also successfully accounts for other known effects to decision making literature, such as: attribute spacing effect \citep{cooke1998multiattribute}; centrality effect \citep{brown2011decision}; background contrast effect \citep{tverskySimonson93}; endowment effect \citep{knetsch1989endowment} and others\footnote{For the full list of the effects MDbS can account for please refer to Noguchi \citeyearonly{noguchi2018multialternative}.}. Overall, authors claim that MDbS can theoretically account for up to 25 context effect variations.

@@

Because of MDbS nature of comparing dimensions with one another, for the purposes of this study, I can not use variables that do not differ across the menus. Additionally, I can not utilize horizontal attributes in my analysis because they do not follow the standard ``greater the better'' mathematical approach to comparison either. However, these variables are used later in this dissertation in chapters \ref{chapter:jmrPaper} and \ref{chapter:hicssPaper}. As a result, I am bound to utilize solely four vertical attributes in my analysis. Table \ref{tab:descriptiveStats} provides descriptive information about variables. I have also multiplied all four vertical variables by \-1 to convert them to negative scale because of MDbS nature of comparing absolute values and the assumption that consumers prefer lower values along vertical dimensions.

@@

In the course of this study, I also introduce an additional dataset sourced from a controlled experiment conducted by Noguchi \citeyearonly{noguchi2018multialternative}, distinct from the primary observational data. Albeit not as diverse in terms of alternatives, dimensions and choice sets, this experimental data carries significant value, not as a principal analytic focus, but rather as a mean to corroborate my main findings. I will apply the same analytical techniques employed in the observational analysis, and utilize experimental dataset as a robustness check to verify the validity of my results. Henceforth, the role of this data is primarily confirmatory.

This data comes from an experiment conducted by Noguchi \citeyearonly{noguchi2018multialternative} where 503 participants, aged 18 to 75, that took part in Amazon Mechanical Turk resulting in 5295 observations total. Participants faced eight randomly sampled decision scenarios with descriptions that consisted of two and three alternative sets, each ternary choice set containing only one of the attraction, compromise, similarity effects. For ternary choice sets, to create a context effect one alternative was randomly selected as focus and third alternative was generated following three scenarios: a) for attraction effect scenario, third options' both dimensions were reduced by 25\% of the difference between the remaining two options' dimensions; b) for compromise scenario, the third option was generated in a way that it would make the randomly chosen target a compromise; c) in similarity scenario 2\% of the difference was added to one dimension while 2\% was subtracted from another dimension for the third option. The dimensions of alternatives are described in the table \ref{tab:noguchiDescriptions}.

@@

After discussing the importance of optimal parameter search and defining the optimization algorithm, the question remaining is the evaluation metric of the MDbS. Previous studies which have applied various dynamic choice models to experimental data, have used the mean absolute error of aggregate choice shares for the entire dataset as main metric. Albeit an interesting approach itself, this will not be a feasible approach for me because the experimental data these models have been applied to entailed ternary choice sets, whereas the observational data is not ternary. It comprised of choice sets with minimum of 2 and maximum of 99 alternatives.

Choice set designers and engineers have long used ``Top n'' accuracy metrics when designing choice sets, or testing the performance of statistical models \citep{ricci2015recommender}. ``Top $n$'' accuracy metric posits whether or not, true class of the option matches the top $n$ predictions of the model. I will follow and adopt this metric because it is well established in the literature, mirrors the real-world decision making and it fits the contribution of the thesis the best. I will use ``Top 1'' accuracy which ranges from 0 to 1 as my metric of optimization for an individual choice set. Because choice sets in the data vary significantly in size, I will use average Top 1 accuracy metric weighted by menu size. This will ensure that smaller menus contribute proportional to their sizes. Also, to comply with differential evolution algorithm's aim of minimization, I will multiply this metric by -1. As an additional measure to explore the parameter space thoroughly, I employ multiple-run approach for differential evolution algorithm. Specifically, I will execute it ten times across the entire dataset. This reposition will allow me to further explore the parameter space, mitigating the risk of missing any region that can potentially contain optimal solution.

@@

Recall that variable ``MDbS output'' refers to the probabilities produced by MDbS. Model 2 results show positive and statistically significant effect for the information provided. It shows that MDbS is able to capture additional information about the choice by accounting for context effects. To better understand the significance of this result, figure \ref{fig:marginsAmadeusGraph} shows the average marginal effects of the information provided by computational model. One can immediately observe the downward trend. It is not surprising. As the number of alternatives increases, each additional alternative adds less to the likelihood of choice than the one before. In the context of menus, it could imply that when there are fewer options (smaller menus), the likelihood of any particular choice being selected is more significantly influenced by MDbS output. On average, for every 0.5 increase in MDbS output the probability of choice has increased by 0.037 percentage points \footnote{Marginal effect of MDbS output across the whole dataset was 0.074.}. This effect was as high as 0.12 percentage points for menus containing as small as 5 alternatives.

@@

At first sight, the results from optimization indicate that the optimal parameters differ across the datasets. Considering the nature of these two datasets, such result is to be expected. While optimal $\alpha$ tends to be higher than the one in observational data, both optimal $\beta_0$ and $\beta_1$ values are lower than their counterparts. Higher $\alpha$ can indicate that for smaller choice sets, MDbS tends to  be less strict in comparison criteria. While the performance metrics seem higher than for the observational data, menus are considerably smaller. Table \ref{tab:optimizationNoghuchiResults} entails further information.

@@

The use of MDbS in current study has shown it has ability to potentially capture wide range of context effects, including attraction, compromise and similarity. While these results have yielded valuable insights, the general nature of MDbS and other computational models is their inability to successfully isolate those effects from one another. Main reason for that is that they have been only tested in experiments with one effect present at a time. When the number of alternatives in the dataset increases the potential interplay between options and the existence of other context effects come into play.

My findings provide a strong foundation which leads to a crucial but also challenging future direction: the development of a methodology which would allow to disentangle this ``general'' context effect. I will computationally differentiate among three main components, attraction, similarity and compromise in a multi-dimensional, multi-alternative choice setting. This goal provides a great motivation for the next chapter of my thesis.

@@

Considering the rapid technological developments of the recent years, I may state with confidence that user control in recommender systems is about to take a huge leap forward and open even more research directions.
